``Interestingness'' in Mathematical Theory Exploration

### 2019-10-04 PhDThesis/clustering.tex
Incorporate feedback

To improve the performance of theory exploration tools, we need to avoid
brute-force search in favour of a more informed strategy. This, in turn,
requires more information about the terms in the given signature. One
generally available source of information is the syntax of a term's definition.

Unfortunately these definitions are over-complicated due to the relatively large
syntax of Haskell, which includes many convenience features (``syntactic
sugar'') and alternative ways to express the same operation (e.g. binding using
\hs{let} or \hs{where}, pattern matching with \hs{if} or \hs{case}, function
abstraction via \hs{f x = ...} or \hs{f = \x -> ...}, etc.). This obscures the
simplicity of the lambda calculus underlying the language, and more importantly
it makes syntactic analysis less useful for our purposes by drawing distinctions
between semantically identical terms. Since theory exploration is concerned only
with semantics (e.g. telling us how code actually behaves), we perform our
analyses on the syntax of the \emph{GHC Core} language instead.

\section{GHC Core}\label{sec:core}
\label{sec:ghccore}

@@

  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote
    repetition and grouping, respectively. Other forms of literal, like machine
    words, are treated in a similar way and have been omitted for brevity.}
  \label{fig:coresyntax}
\end{figure}

@@

\begin{figure}
  \begin{subfigure}[haskell]{\textwidth}
    \underline{\texttt{Haskell definitions}}
    \begin{haskell}
plus    Z  y = y
plus (S x) y = S      (plus x y)

@@

even    Z  = True
even (S n) = odd n
    \end{haskell}
    \smallskip
  \end{subfigure}
  \begin{subfigure}[plus]{\textwidth}
    \begin{small}
      \underline{\texttt{plus}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       (Alt (DataAlt "Z") (Var (Local "y")))
                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
                                               (App (App (Var (Global "plus"))
                                                         (Var (Local  "x")))
                                                    (Var (Local "y"))))
                                          "x")))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[mult]{\textwidth}
    \begin{small}
      \underline{\texttt{mult}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
                                                    (Var (Local  "y")))
                                               (App (App (Var (Global "mult"))
                                                         (Var (Local  "x")))
                                                    (Var (Local  "y"))))
                                          "x")))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[odd]{\textwidth}
    \begin{small}
      \underline{\texttt{odd}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "False")))
              (Alt (DataAlt "S") (App (Var (Global "even"))
                                      (Var (Local  "n")))
                                 "n"))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[even]{\textwidth}
    \begin{small}
      \underline{\texttt{even}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "True")))
              (Alt (DataAlt "S") (App (Var (Global "odd"))
                                      (Var (Local  "n")))
                                 "n"))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \caption{Example Haskell functions and their translations into the Core syntax
    of Figure \ref{fig:coresyntax}. Notice the introduction of explicit
    $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to
    represent piecewise definitions. Fresh variables are chosen arbitrarily
    as \texttt{"a"}, \texttt{"b"}, etc.}
  \label{fig:coreexample}
\end{figure}

@@

As a further simplification, we also replace data constructors in definitions
with normal functions, e.g. replacing occurrences of a constructor
\hs{S :: Nat -> Nat} with a function \hs{s} defined as \hs{s x = S x} (this is
known as the $\eta$-expansion of \hs{S}). We do this to remove more distinctions
from the syntax, and also to reduce cross-language differences (in particular,
the Isabelle language used by the \isascheme{}, \isacosy{} and \hipster{} theory
exploration systems treats constructors in a different way to Haskell, so having
constructors in our data would complicate comparisons).

\section{Feature Extraction}

@@

The process of turning raw data into feature vectors is called \emph{feature
  extraction}, and we adapt the methodology of \emph{recurrent clustering}
proposed in~\cite{DBLP:journals/corr/HerasK14,heras2013proof}, since this
handles recursive tree structures, resolves references and produces numeric
values. We suggest a new recurrent clustering and feature extraction algorithm
for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity
and differences to the existing approaches of \textsc{ML4PG} and
\textsc{ACL2(ml)}.

We describe our algorithm in two stages: the first transforms the nested
structure of expressions into a flat vector representation; the second converts
the discrete symbols of Core syntax into features (real numbers), which we will
denote using the function $\phi$. Before introducing our algorithm, we briefly
describe \emph{clustering}, and its use in \emph{recurrent clustering}.

\subsection{Clustering}

Clustering is an unsupervised machine learning task for grouping $n$ data points
using a similarity metric. There are many variations on this theme, but in our
case we make the following choices:

\begin{itemize}
\item For simplicity, we use the ``rule of thumb'' given in
  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
\item Data points will be $d$-dimensional feature vectors, as defined above.
\item We will use euclidean distance (denoted $e$) as our similarity metric.
\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
  \cite{lloyd1982least}.
\end{itemize}

We choose k-means because it is a standard approach and we are more concerned
with the application of feature extraction to Haskell and its use in theory
exploration, rather than precise tuning of learning algorithms. There are many
other clustering algorithms we could use, such as \emph{expectation
  maximisation}, but experiments with \textsc{ML4PG} have shown little
difference in their results; it seems that the quality of the features is the
bottleneck to learning, so there is no reason to avoid a fast algorithm like
k-means. We use a Haskell implementation provided by the package
\hPackage{k-means}.

Since k-means is iterative, we will use function notation to denote time steps,
so $x(t)$ denotes the value of $x$ at time $t$. We denote the clusters as $C^1$
to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:

Before k-means starts, we must choose \emph{seed} values for
$\vect{m}^i(0)$. Many methods have been proposed for choosing these values
\cite{arthur2007k}. For simplicity, we will choose values randomly from our
data set $S$; this is known as the Forgy method.

The elements of each cluster $C^i(t)$ are those data points closest to the mean
value at the previous time step:

Many conservative improvements can be made to the standard k-means algorithm
described above. For example, a more efficient approach like
\emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make larger input sizes
more practical to cluster; the \emph{k-means++}
approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
select the ``seed'' values for the first timestep; and the \emph{x-means}
algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use
(instead of relying on our ``rule of thumb''). However, like the choice of
k-means versus other clustering algorithms, these algorithmic improvements would
need higher quality data than we are capable of producing in these early
experiments.

\emph{Recurrent clustering} has been implemented in the
ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
\textsc{ACL2(ml)} for ACL2~\cite{heras2013proof}. Both approaches transform
syntax trees into fixed-size matrices, then concatenate the rows of these
matrices to produce feature vectors. \textsc{ACL2(ml)} produces matrices by
\emph{summarising} information about the tree; for example, one column counts
the number of variables appearing at each tree level, others count the number of
function symbols which are nullary, unary, binary, etc. In contrast, the
algorithm of ML4PG converts tokens of the syntax tree directly into numbers,
arranging them in the matrix based on their position in the syntax tree. It is
this latter algorithm that we adapt for analysing GHC Core expressions.

Recurrent clustering solves this dichotomy by introducing a finer granularity to
the treatment of identifiers, by first clustering the definitions that they
refer to. If two definitions appear in the same cluster, their identifiers will
be denoted by similar numbers during subsequent feature extraction; if they
appear in different clusters, their numbers will differ more during feature
extraction. It is this use of clustering to determine features that makes the
approach \emph{recurrecnt}.

\subsection{Expressions to Vectors}
\label{sec:expressionstovectors}

The first step of our feature extraction algorithm is to convert the tree
structure of Core expressions to a flat vector. Since k-means clustering
considers the elements of a feature vector to be \emph{orthogonal}, we must
ensure that similar expressions not only give rise to similar numerical values,
but crucially that these values appear \emph{at the same position} in the
feature vectors. Different patterns of nesting can alter the ``shape'' of
expressions, so simple traversals (breadth-first, depth-first, post-order, etc.)
may cause features from equivalent sub-expressions to be mis-aligned. For
example, consider the following Core expressions, which represent pattern-match
clauses with different patterns but the same body (\hs{\vlocal{y}}):

\begin{equation}\label{eq:xy}
  \begin{array}{r@{}l@{}l@{}}
    X\ &=\ \CAlt\ (\CDataAlt\ \id{C})\ & (\vlocal{y}) \\
    Y\ &=\ \CAlt\ \CDefault\           & (\vlocal{y})
  \end{array}
\end{equation}

@@

\begin{small}
  \begin{equation}
    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
      breadthFirst(X)\ &=\ (\feature{\CAlt},\
                           &\feature{\CDataAlt},\
                           &\feature{\CVar},\
                           &\feature{\id{C}},\
                           &\feature{\CLocal},\
                           &\feature{\id{y}} &) \\
      breadthFirst(Y)\ &=\ (\feature{\CAlt},\
                           &\feature{\CDefault},\
                           &\feature{\CVar},\
                           &\feature{\CLocal},\
                           &\feature{\id{y}},\
                           &0 &)
    \end{array}
  \end{equation}
\end{small}

@@

\begin{small}
  \begin{equation}\label{eq:flattened}
    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
      featureVec(X)\ &=\ (\feature{\CAlt},\
                         &0,\
                         &\feature{\CDataAlt},\
                         &\feature{\CVar},\
                         &\feature{\id{C}},\
                         &\feature{\CLocal},\
                         &\feature{\id{y}},\
                         &0 &) \\
      featureVec(Y)\ &=\ (\feature{\CAlt},\
                         &0,\
                         &\feature{\CDefault},\
                         &\feature{\CVar},\
                         &\feature{\CLocal},\
                         &0,\
                         &\feature{\id{y}},\
                         &0 &)
    \end{array}
  \end{equation}
\end{small}

@@

\begin{figure}
  \centering
  \begin{scriptsize}
      \Tree[ .$\feature{\CLam}$
                $\feature{\id{a}}$
                [ .$\feature{\CCase}$
                     [ .$\feature{\CVar}$
                          [ .$\feature{\CLocal}$
                               $\feature{\id{a}}$ ]]
                     $\feature{\id{b}}$
                     [ .$\feature{\CAlt}$
                          $\feature{\CDataAlt}$
                          [ .$\feature{\CVar}$
                               $\feature{\CConstructor}$ ]]
                     [ .$\feature{\CAlt}$
                          $\feature{\CDataAlt}$
                          [ .$\feature{\CApp}$
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CGlobal}$
                                         $\feature{\id{even}}$ ]]
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CLocal}$
                                         $\feature{\id{n}}$ ]]]
                          $\feature{\id{n}}$ ]]]
  \end{scriptsize}
  \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the
    expression \hs{odd} from Figure \ref{fig:coreexample}. Each (sub-) rose tree
    is rendered with its feature at the node and sub-trees beneath.}
\end{figure}

@@

\begin{figure}
    \begin{equation*}
      \begin{bmatrix}
        \feature{\CLam}      & 0                       & 0                 & 0                   & 0               & 0                \\
        \feature{\id{a}}     & \feature{\CCase}        & 0                 & 0                   & 0               & 0                \\
        \feature{\CVar}      & \feature{\id{b}}        & \feature{\CAlt}   & \feature{\CAlt}     & 0               & 0                \\
        \feature{\CLocal}    & \feature{\CDataAlt}     & \feature{\CVar}   & \feature{\CDataAlt} & \feature{\CApp} & \feature{\id{n}} \\
        \feature{\id{a}}     & \feature{\CConstructor} & \feature{\CVar}   & \feature{\CVar}     & 0               & 0                \\
        \feature{\CGlobal}   & \feature{\CLocal}       & 0                 & 0                   & 0               & 0                \\
        \feature{\id{even}}  & \feature{\id{n}}        & 0                 & 0                   & 0               & 0
      \end{bmatrix}
    \end{equation*}
    \caption{Matrix generated from Figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
    \label{fig:matrixexample}
\end{figure}

@@

We now define the function $\phi$, which turns terminal symbols of Core syntax
into features (real numbers). For known language features, such as
$\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
assign a value to each, in a similar way to~\cite{DBLP:journals/corr/HerasK14}
in Coq. We use a constant $\alpha$ to separate these values from those of other
tokens (e.g. identifiers), but the order is essentially
arbitrary:~\footnote{In~\cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina
  tokens like \coq{fix} and \coq{cofix} are grouped together to reduce
  redundancy; we do not group tokens, but we do put ``similar'' tokens close
  together, such as \CLocal\ and \CGlobal.}

@@

Instead, we use the approach taken nu ML4PG and encode global identifiers
\emph{indirectly}, by looking up the expressions which they \emph{reference}:

\begin{equation} \label{eq:globalfeature}
  \feature{g \in \mathcal{G}} =
    \begin{cases}
      i + 3 \alpha \quad & \text{if $g \in C_i$} \\
      f_{recursion}         & \text{otherwise}
    \end{cases}
\end{equation}

Where $C_i$ are our clusters (in some arbitrary order). This is where the
recurrent nature of the algorithm appears: to determine the contents of $C_i$,
we must perform k-means clustering \emph{during} feature extraction; yet that
clustering step, in turn, requires that we perform feature extraction.

For this recursive process to be well-founded, we perform a topological sort on
declarations based on their dependencies (the expressions they reference). In
this way, we can avoid looking up expressions which haven't been clustered
yet. To handle mutual recursion we use Tarjan's algorithm~\cite{tarjan1972depth}
to produce a sorted list of \emph{strongly connected components} (SCCs), where
each SCC is a mutually-recursive sub-set of the declarations. If an identifier
cannot be found amongst those clustered so far, it must appear in the same SCC
as the expression we are processing; hence we give that identifier the constant
feature value $f_{recursion}$.

@@

Even in this very simple example we can see a few features of our algorithm
emerge. For example, \hs{odd} and \hs{even} will always appear in the same
cluster, since they only differ in their choice of constructor names (which are
discarded by $toTree$) and recursive calls (which are replaced by
$f_{recursion}$).

@@

We provide an implementation of our recurrent clustering algorithm in a tool
called \mlforhs{}. We obtain Core ASTs from Haskell packages using a plugin
for the GHC compiler~\footnote{Available at
  \url{https://github.com/warbo/ast-plugin}}, which emits a serialised version
of each Core definition as it is being compiled. This approach is more robust
than, for example, parsing source files, since it avoids the complications of
preprocessors and language extensions altering the syntax.

A post-processing stage determines which Core definitions can be used by
\quickspec{}, based on their type, visibility (whether they are encapsulated inside
their module or visible to importers), etc. For each definition, we simply use
the GHCi interpreter to check if calling \quickspec{} with that argument is
well-typed. This information, along with type signatures, arity, etc. are stored
alongside the Core definitions in JSON format.

>>>>

% TODO: Say what's in this chapter
In this chapter we consider how the contents of a given signature can be
analysed, such that subsequent exploration can be performed more efficiently. We
frame this analysis as a task of \emph{unsupervised machine learning}, and
describe how well-known \emph{clustering algorithms} can utilise the syntax of
definitions to determine their similarity. This requires a detailed description
of Haskell definition syntax, in particular of the ``Core'' intermediate
language of the GHC compiler. We identify the difficulty posed by references,
and how these can be made less opaque through the use of \emph{recurrent
  clustering}.

\section{Syntax}

To improve the performance of theory exploration tools, we need to avoid
brute-force search in favour of a more informed strategy.
\iffalse
TODO: Alison: This makes it sound as though current TE tools all use brute force
(whereas all the tools I know already use informed strategies).
MY THOUGHTS: ``more informed'' is a bit of weasel-wording from me; implying that
they're not informed and that we're making them informed; whereas they're
actually already somewhat informed, and we're trying to add to that. I agree
that this weasel-wording should be replaced!
\fi
This, in turn,
requires more information \iffalse TODO: Alison: More than what? \fi about the
terms in the given signature. One
generally available source of information is the syntax of a term's definition.

Unfortunately these definitions are over-complicated due to the relatively large
syntax of Haskell, which includes many convenience features (``syntactic
sugar'') and alternative ways to express the same operation (e.g. binding using
\hs{let} or \hs{where}, pattern matching with \hs{if} or \hs{case}, function
abstraction via \hs{f x = ...} or \hs{f = \x -> ...}, etc.). This obscures the
simplicity of the lambda calculus underlying the language, and more importantly
it makes syntactic analysis less useful for our purposes by drawing distinctions
between semantically identical terms. Since theory exploration is concerned only
with semantics (telling us how code actually behaves), we perform our analyses
on the simpler syntax of the \emph{GHC Core} language instead.

\subsection{GHC Core}\label{sec:core}
\label{sec:ghccore}

@@

  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote
    repetition and grouping, respectively \iffalse TODO: Alison: Include examples? \fi. Other forms of literal, like machine
    words, are treated in a similar way and have been omitted for brevity.}
  \label{fig:coresyntax}
\end{figure}

@@

\begin{figure}
  \begin{subfigure}[haskell]{\textwidth}
    \underline{\texttt{Haskell definitions}}
    \iffalse
    TODO: Alison: Big space
    MY THOUGHTS: I wanted to align things more neatly...
    \fi
    \begin{haskell}
plus    Z  y = y
plus (S x) y = S      (plus x y)

@@

even    Z  = True
even (S n) = odd n
    \end{haskell}
    \smallskip
  \end{subfigure}
  \begin{subfigure}[plus]{\textwidth}
    \begin{small}
      \underline{\texttt{plus}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       (Alt (DataAlt "Z") (Var (Local "y")))
                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
                                               (App (App (Var (Global "plus"))
                                                         (Var (Local  "x")))
                                                    (Var (Local "y"))))
                                          "x")))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[mult]{\textwidth}
    \begin{small}
      \underline{\texttt{mult}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
                                                    (Var (Local  "y")))
                                               (App (App (Var (Global "mult"))
                                                         (Var (Local  "x")))
                                                    (Var (Local  "y"))))
                                          "x")))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[odd]{\textwidth}
    \begin{small}
      \underline{\texttt{odd}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "False")))
              (Alt (DataAlt "S") (App (Var (Global "even"))
                                      (Var (Local  "n")))
                                 "n"))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[even]{\textwidth}
    \begin{small}
      \underline{\texttt{even}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "True")))
              (Alt (DataAlt "S") (App (Var (Global "odd"))
                                      (Var (Local  "n")))
                                 "n"))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \caption{Example Haskell functions and their translations into the Core syntax
    of Figure~\ref{fig:coresyntax}. Notice the introduction of explicit
    $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to
    represent piecewise definitions. Fresh variables are chosen arbitrarily
    as \texttt{"a"}, \texttt{"b"}, etc.}
  \label{fig:coreexample}
\end{figure}

@@

Note that in test data such as our Theory Exploration Benchmark, data
constructors in definitions are replaced with normal functions, e.g. replacing
occurrences of a constructor \hs{S :: Nat -> Nat} with a function \hs{s} defined
as \hs{s x = S x} (this is known as the $\eta$-expansion of \hs{S}). This
removes even more distinctions from the syntax, and also reduces
cross-language differences (in particular, the Isabelle language used by the
\isascheme{}, \isacosy{} and \hipster{} theory exploration systems treats
constructors in a different way to Haskell, so removing constructors from our
data simplifies cross-language comparisons).

\section{Feature Extraction}
\label{sec:featureextraction}

@@

The process of turning raw data into feature vectors is called \emph{feature
  extraction}, and we adapt the methodology of \emph{recurrent clustering}
proposed in~\cite{DBLP:journals/corr/HerasK14,heras2013proof}, since this
handles recursive tree structures, resolves references and produces numeric
values. We suggest a new recurrent clustering and feature extraction algorithm
for Haskell (shown in Algorithm~\ref{alg:recurrent}) and compare its similarity
and differences to the existing approaches of \textsc{ML4PG} and
\textsc{ACL2(ml)}.

We describe our algorithm in two stages: the first transforms the nested
structure of expressions into a flat vector representation; the second converts
the discrete symbols of Core syntax into features (real numbers), which we will
denote using the function $\phi$.

\subsubsection{Expressions to Vectors}
\label{sec:expressionstovectors}

The first step of our feature extraction algorithm is to convert the tree
structure of Core expressions to a flat vector. Since k-means clustering
considers the elements of a feature vector to be \emph{orthogonal}, we must
ensure that similar expressions not only give rise to similar numerical values,
but crucially that these values appear \emph{at the same position} in the
feature vectors. Different patterns of nesting can alter the ``shape'' of
expressions, so simple traversals (breadth-first, depth-first, post-order, etc.)
may cause features from equivalent sub-expressions to be mis-aligned. For
example, consider the following Core expressions, which represent pattern-match
clauses with different patterns but the same body (\vlocal{y}):

\begin{equation*}\label{eq:xy}
  \begin{array}{r@{}l@{}l@{}}
    X\ &=\ \CAlt\ (\CDataAlt\ \id{C})\ & (\vlocal{y}) \\
    Y\ &=\ \CAlt\ \CDefault\           & (\vlocal{y})
  \end{array}
\end{equation*}

@@

\begin{small}
  \begin{equation*}
    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
      breadthFirst(X)\ &=\ (\feature{\CAlt},\
                           \ &\feature{\CDataAlt},\
                           \ &\feature{\CVar},\
                           \ &\feature{\id{C}},\
                           \ &\feature{\CLocal},\
                           \ &\feature{\id{y}} &) \\
      breadthFirst(Y)\ &=\ (\feature{\CAlt},\
                           \ &\feature{\CDefault},\
                           \ &\feature{\CVar},\
                           \ &\feature{\CLocal},\
                           \ &\feature{\id{y}},\
                           \ &0 &)
    \end{array}
  \end{equation*}
\end{small}

@@

\begin{small}
  \begin{equation*}\label{eq:flattened}
    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
      featureVec(X)\ &=\ (\feature{\CAlt},\
                         &0,\
                         &\feature{\CDataAlt},\
                         &\feature{\CVar},\
                         &\feature{\id{C}},\
                         &\feature{\CLocal},\
                         &\feature{\id{y}},\
                         &0 &) \\
      featureVec(Y)\ &=\ (\feature{\CAlt},\
                         &0,\
                         &\feature{\CDefault},\
                         &\feature{\CVar},\
                         &\feature{\CLocal},\
                         &0,\
                         &\feature{\id{y}},\
                         &0 &)
    \end{array}
  \end{equation*}
\end{small}

@@

\begin{figure}
  \centering
  \begin{scriptsize}
      \Tree[ .$\feature{\CLam}$
                $\feature{\id{a}}$
                [ .$\feature{\CCase}$
                     [ .$\feature{\CVar}$
                          [ .$\feature{\CLocal}$
                               $\feature{\id{a}}$ ]]
                     $\feature{\id{b}}$
                     [ .$\feature{\CAlt}$
                          $\feature{\CDataAlt}$
                          [ .$\feature{\CVar}$
                               $\feature{\CConstructor}$ ]]
                     [ .$\feature{\CAlt}$
                          $\feature{\CDataAlt}$
                          [ .$\feature{\CApp}$
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CGlobal}$
                                         $\feature{\id{even}}$ ]]
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CLocal}$
                                         $\feature{\id{n}}$ ]]]
                          $\feature{\id{n}}$ ]]]
  \end{scriptsize}
  \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the
    expression \hs{odd} from Figure~\ref{fig:coreexample}. Each (sub-) rose tree
    is rendered with its feature at the node and sub-trees beneath.}
\end{figure}

@@

\begin{figure}
    \begin{equation*}
      \begin{bmatrix}
        \feature{\CLam}      & 0                       & 0                 & 0                   & 0               & 0                \\
        \feature{\id{a}}     & \feature{\CCase}        & 0                 & 0                   & 0               & 0                \\
        \feature{\CVar}      & \feature{\id{b}}        & \feature{\CAlt}   & \feature{\CAlt}     & 0               & 0                \\
        \feature{\CLocal}    & \feature{\CDataAlt}     & \feature{\CVar}   & \feature{\CDataAlt} & \feature{\CApp} & \feature{\id{n}} \\
        \feature{\id{a}}     & \feature{\CConstructor} & \feature{\CVar}   & \feature{\CVar}     & 0               & 0                \\
        \feature{\CGlobal}   & \feature{\CLocal}       & 0                 & 0                   & 0               & 0                \\
        \feature{\id{even}}  & \feature{\id{n}}        & 0                 & 0                   & 0               & 0
      \end{bmatrix}
    \end{equation*}
    \caption{Matrix generated from Figure~\ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
    \label{fig:matrixexample}
\end{figure}

@@

We now define the function $\phi$, which turns terminal symbols of Core syntax
into features (real numbers). For known language features, such as
$\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
assign a value to each, in a similar way to~\cite{DBLP:journals/corr/HerasK14}
in Coq. We use a constant $\alpha$ to separate these values from those of other
tokens (e.g. identifiers), but the order is essentially
arbitrary~\footnote{In~\cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina
  tokens like \coq{fix} and \coq{cofix} are grouped together to reduce
  redundancy; we do not group tokens, but we do put ``similar'' tokens close
  together, such as \CLocal\ and \CGlobal.}:

@@

Instead, we use the approach taken by ML4PG and encode global identifiers
\emph{indirectly}, by looking up the expressions which they \emph{reference}:

\begin{equation} \label{eq:globalfeature}
  \feature{g \in \mathcal{G}} =
    \begin{cases}
      i + 3 \alpha \quad & \text{if $g \in C^i$} \\
      f_{recursion}         & \text{otherwise}
    \end{cases}
\end{equation}

Here $f_{recursion}$ is a constant value which is used for recursive calls, like
those in the \hs{plus} and \hs{mult} definitions above. It will also appear in
cases of mutual recursion, like the call to \hs{odd} within the definition of
\hs{even} and \emph{vice versa}.

$C$ contains the results (in some arbitrary order) of a \emph{clustering
  algorithm} (described below in \S~\ref{sec:clustering}) applied to all of
those definitions which are referenced non-recursively by this one. When we
encounter such a reference, we look up which cluster that definition appeared
in (denoted $C^i$), and use that index ($i$) as our feature value (using
$\alpha$ to offset from other features as before).

Defining the \emph{input} (feature vectors) for one definition in terms of the
\emph{output} (clusters) of some other definitions gives rise to a
\emph{recurrent clustering} algorithm, and blurs the boundary between feature
extraction and clustering, which are normally distinct steps.

\section{Clustering}
\label{sec:clustering}

Clustering is an unsupervised machine learning task for dividing $n$ data points
into $k$ groups (``clusters'') such that points deemed similar a given metric
tend to be grouped together. There are many variations on this theme, but in our
case we make the following choices:

\begin{itemize}
\item For simplicity, we use the ``rule of thumb'' given in
  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
\item Data points will be $d$-dimensional feature vectors, as defined above in
  \S~\ref{sec:featureextraction}.
\item We will use euclidean distance (denoted $e$) as our similarity metric.
\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
  \cite{lloyd1982least}.
\end{itemize}

\iffalse
TODO: Alison: It would be good to have justification of the previous
items as well.
\fi

We choose k-means because it is a standard approach \iffalse TODO: Alison: Is
there a reference for this? \fi. There are many other clustering algorithms we
could use, such as \emph{expectation maximisation}, but experiments with
\textsc{ML4PG} have shown little difference in their results; \iffalse TODO:
Alison: Whose experiments? Yours (include something in appendix?) Someone else's
(reference)? MY THOUGHTS: Conversation with Katya; maybe check ML4PG papers, or
omit \fi it seems that \iffalse TODO: Alison: a bit informal and vague \fi the
quality of the features is the bottleneck to learning, so there is no reason to
avoid a fast algorithm like k-means. We use a Haskell implementation provided by
the package \hPackage{k-means}.

Since k-means is iterative, we will use function notation to denote time steps,
so $x(t)$ denotes the value of $x$ at time $t$. \iffalse TODO: Alison: Pick a
tense and stick to it. I'd suggest the present tense throughout \fi We denote
the clusters as $C^1$
to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
which we denote as $\vect{m}^1$ to $\vect{m}^k$. The components of these vectors
are treated as orthogonal, hence the $j$th component of $\vect{m}^i$ is the
average of the $j$th components of all vectors in cluster $C^i$:

Before k-means starts, we must choose \emph{seed} values for
$\vect{m}^i(0)$. Many methods have been proposed for choosing these
values~\cite{arthur2007k}. For simplicity, we will choose values randomly from
our data set $S$; this is known as the Forgy method. \iffalse TODO: Alison: Any
stronger justification? Why are the other methods not appropriate? \fi

The elements of each cluster $C^i(t)$ are those data points closest to the mean
value at the previous time step $\vect{m}^i(t-1)$:

Many conservative improvements can be made to the standard k-means algorithm
described above. For example, a more efficient approach like
\emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make larger input sizes
more practical to cluster; the \emph{k-means++}
approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
select the ``seed'' values for the first timestep; and the \emph{x-means}
algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use
(instead of relying on our ``rule of thumb''). However, like the choice of
k-means versus other clustering algorithms, the benefits of more, higher-quality
data seem greater than those offered by such algorithmic tweaking. \iffalse
TODO: Also describe better reasons for avoiding these improvements, i.e. they
are algorithmically complicated and require multiple passes over the input data
\fi

\emph{Recurrent clustering} has been implemented in the
ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
\textsc{ACL2(ml)} for ACL2~\cite{heras2013proof}. Both approaches transform
syntax trees into fixed-size matrices, then concatenate the rows of these
matrices to produce feature vectors. \textsc{ACL2(ml)} produces matrices by
\emph{summarising} information about the tree; for example, one column counts
the number of variables appearing at each tree level, others count the number of
function symbols which are nullary, unary, binary, etc. In contrast, the
algorithm of ML4PG converts tokens of the syntax tree directly into numbers,
arranging them in the matrix based on their position in the syntax tree. It is
this latter algorithm that we have adapted for analysing GHC Core expressions.

Recurrent clustering solves this dichotomy by introducing a finer granularity to
the treatment of identifiers, by first clustering the definitions that they
refer to. If two definitions appear in the same cluster, references to those
definitions will be denoted by similar numbers during subsequent feature
extraction; if they appear in different clusters, their numbers will differ more
during feature extraction. It is this use of clustering to determine features
that makes the approach \emph{recurrent}.

For this recursive process to be well-founded, we perform a topological sort on
definitions based on their dependencies (the definitions they reference). In
this way, we can avoid looking up definitions which haven't been clustered
yet. To handle mutual recursion we use Tarjan's algorithm~\cite{tarjan1972depth}
to produce a sorted list of \emph{strongly connected components} (SCCs), where
each SCC is a mutually-recursive sub-set of the declarations. If an identifier
cannot be found amongst those clustered so far, it must appear in the same SCC
as the expression we are processing; hence we give that identifier a constant
feature value denoted as $f_{recursion}$.

@@

Even in this very simple example we can see a few features of our algorithm
emerge. For example, \hs{odd} and \hs{even} will always appear in the same
cluster, since they only differ in their choice of constructor names (which are
discarded by $toTree$) and recursive calls (which are both replaced by
$f_{recursion}$).

@@

We provide an implementation of our recurrent clustering algorithm in a tool
called \mlforhs{}~\footnote{Available at
  \url{https://github.com/warbo/ml4hsfe}}. We obtain Core ASTs from Haskell
packages using a plugin for the GHC compiler~\footnote{Available at
  \url{https://github.com/warbo/ast-plugin}}, which emits a serialised version
of each Core definition as it is being compiled. This approach is more robust
than, for example, parsing source files, since it avoids the complications of
preprocessors and language extensions altering the syntax.

A post-processing stage determines which Core definitions are suitable for
exploration, based on their visibility (whether they are encapsulated inside
their module or visible to importers), whether a data generator is available,
etc. This information, along with type signatures, arity, etc. are stored
alongside the Core definitions in JSON format.
